---
title: "Public Policy 760 Final Project"
author: "Theodore Landsman"
date: "12/2/2020"
output: html_document
---


### Data Import

```{r, warning = FALSE, message=FALSE }
library(tidyverse)
library(readxl)
library(smacof)
library(ggplot2)

wyoming <- read_csv("~/Documents/Wyoming data import.csv")
AlaskaCVR <- read_excel("~/Documents/AlaskaCVR.xlsx", na = "under")

names<-colnames(wyoming)
names[6:10]<- c(1,2,3,4,5)
colnames(wyoming) <- names
wyoming_l <- wyoming %>%
  select(BallotID, `1` ,`2` ,`3` ,`4`, `5` ) %>%
  pivot_longer(-BallotID, names_to = 'rank', values_to = "candidate") %>%
  filter(!candidate %in% c('under','over')) %>%
  pivot_wider(names_from = candidate, values_from = rank, values_fn = first)

```

### Data Preprocessing
```{r, warning = FALSE}
w_matrix <- wyoming_l[,2:10]
# Impute missing values
w_mat <- w_matrix %>%
    mutate_all(~replace(., is.na(.), 6))
w_mat <- mapply(w_mat, FUN=as.numeric)
principle_components <- w_mat%>%prcomp()
w_pca <-bind_cols(wyoming_l,as_tibble(principle_components$x))
```

### Cluster Analysis
```{r, warning=FALSE}
k_pca<-w_pca
library(broom)
# predict four clusters
w_kmeans <-kmeans(w_mat,centers = 4,# number of clusters
                  nstart = 100)# number of random starts

k_pca$cluster<-w_kmeans$cluster
```

It appears that Biden supports (1) are overwhelmingly in cluster 3, while Sanders supporters are largely in Cluster 2. Cluster 4 seems to be voters who liked both Warren but also Biden and Sanders. Cluster 1 looks to be dominated by Biden supporters who also liked some of the more eccentric also-rans like Bloomberg and Steyer. 

```{r}
ggplot(data = k_pca) +
  geom_point(aes(x = PC1, y =PC2, color = factor(cluster)),alpha = .4) + 
  theme_minimal()
```

### With three clusters
```{r}
k_pca<-w_pca
# predict four clusters
w_kmeans <-kmeans(w_mat,centers = 3,# number of clusters
                  nstart = 100)# number of random starts

k_pca$cluster<-w_kmeans$cluster
ggplot(data = k_pca) +
  geom_point(aes(x = PC1, y =PC2, color = factor(cluster)),alpha = .4) + 
  theme_minimal()
```

With three clusters, Cluster 3 is the Biden cluster, Cluster 1 is a combination of Biden and Sanders voters, and Cluster 2 is a combination of Sanders and Warren Supporters. Plotting the data, we can see that Clusters 2 and 3 occupy the extremes of the PC1 dimension, while Cluster 1 is more moderate. 

```{r}
## This takes way too long, maybe a project for AWS!
library(factoextra)
#fviz_nbclust(w_mat, FUN = kmeans, method = "wss")
# fviz_nbclust(pca_data, FUN = kmeans, method = "gap_stat") This one crashes because it runs out of memory
#fviz_nbclust(w_mat, FUN = kmeans, method = "silhouette")
```

Optimal cluster pattern analysis does not seem to yield a conclusive result, but 2 seems to be the best in terms of optimizing between them. 

### With two clusters
```{r}
k_pca <- w_pca
# predict four clusters
w_kmeans <-kmeans(w_mat,centers = 2,# number of clusters
                  nstart = 100)# number of random starts


k_pca$cluster<-w_kmeans$cluster

ggplot(data = k_pca) +
  geom_point(aes(x = PC1, y =PC2, color = factor(cluster)),alpha = .4) + 
  theme_minimal()
ggplot(data = k_pca) +
  geom_density(aes(x = PC1, fill = factor(cluster)),alpha = .4) + 
  theme_minimal()

```

Cluster 2, Biden Coalition, Cluster 1, numerically much smaller Sanders-Warren coalition. 